---
title: 'Weather Data Analysis'
output:
  html_document:
    css: styles.css
    includes:
      in_header: header.html
date: "`r Sys.Date()`"
---

## Introduction

In this report, we analyze weather data collected from London, including temperature, wind speed, and tide heights. The data is continuously and automatically updated via a Python-based data pipeline, ensuring that our analysis reflects the most current weather conditions. Our goal is to explore various aspects of the weather patterns, identify trends, and visualize key metrics to gain a deeper understanding of local weather conditions.

## Data Overview

```{r schema, message=FALSE, warning=FALSE, echo=FALSE}

# For database connection and interaction
library(DBI)
library(RSQLite)

# For data manipulation and cleaning
library(dplyr)

# For creating and styling tables in HTML
library(kableExtra)

# For data visualization
library(ggplot2)
library(corrplot)


# Define the relative path to the database file
db_file_path <- "data/weather_data.db"

# Connect to the SQLite database
con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)

# List all tables
tables <- dbListTables(con)

# Get foreign key relationships for each table
fk_relationships <- lapply(tables, function(table) {
  dbGetQuery(con, paste("PRAGMA foreign_key_list(", table, ");", sep = ""))
})
names(fk_relationships) <- tables

# Create a data frame to store relationships
relationship_data <- do.call(rbind, lapply(names(fk_relationships), function(table) {
  df <- fk_relationships[[table]]
  if (nrow(df) > 0) {
    data.frame(
      Table = table,
      FK_Column = df$from,
      Referenced_Table = df$table,
      Referenced_Column = df$to,
      stringsAsFactors = FALSE
    )
  } else {
    NULL
  }
}))

# Load necessary libraries
library(dplyr)
library(knitr)
library(kableExtra)

# Assuming `relationship_data` is your data frame with foreign key relationships
relationship_data %>%
  kable("html", caption = "Foreign Key Relationships", escape = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE, color = "black") %>%  # 'Table' column
  column_spec(3, bold = TRUE, color = "black") %>%  # 'Referenced Table' column
  column_spec(2, color = "blue") %>%                # 'From Column'
  column_spec(4, color = "blue") %>%                # 'To Column'
  row_spec(0, background = "#f2f2f2")               # Optional: Light grey background for header row
```


```{r clean_data, echo=FALSE, warning=FALSE, message=FALSE}
library(DBI)
library(RSQLite)
library(dplyr)


# Read data from the relevant table
data <- dbReadTable(con, "WeatherReports")

# Read lookup tables
visibility_levels <- dbReadTable(con, "VisibilityLevels")
weather_conditions <- dbReadTable(con, "WeatherConditions")
wind_directions <- dbReadTable(con, "WindDirections")
pollution_levels <- dbReadTable(con, "PollutionLevels")
pollen_levels <- dbReadTable(con, "PollenLevels")
uv_index_levels <- dbReadTable(con, "UVIndexLevels")

# Clean numeric columns
data <- data %>%
  mutate(
    high_temperature = as.numeric(gsub("[^0-9.]", "", as.character(high_temperature))),
    low_temperature = as.numeric(gsub("[^0-9.]", "", as.character(low_temperature))),
    current_temperature = suppressWarnings(as.numeric(gsub("[^0-9.]", "", as.character(current_temperature)))),
    wind_speed = suppressWarnings(as.numeric(gsub("[^0-9.]", "", as.character(wind_speed)))),
    humidity = suppressWarnings(as.numeric(gsub("[^0-9.]", "", as.character(humidity)))),
    chance_of_precipitation = suppressWarnings(as.numeric(gsub("[^0-9.]", "", as.character(chance_of_precipitation)))),
    pressure = suppressWarnings(as.numeric(gsub("[^0-9.]", "", as.character(pressure))))
  )

# Convert time_of_search to POSIXct
data$time_of_search <- as.POSIXct(data$time_of_search, format = "%Y-%m-%d %H:%M")

# Extract date from time_of_search
data$date <- as.Date(data$time_of_search)

# Keep time columns as text
data$sunset <- as.character(data$sunset)
data$sunrise <- as.character(data$sunrise)
data$low_tide_morning_time <- as.character(data$low_tide_morning_time)
data$high_tide_morning_time <- as.character(data$high_tide_morning_time)
data$low_tide_evening_time <- as.character(data$low_tide_evening_time)
data$high_tide_evening_time <- as.character(data$high_tide_evening_time)

# Join with lookup tables
data <- data %>%
  left_join(visibility_levels, by = c("visibility_id" = "visibility_id")) %>%
  rename(visibility_description = description) %>%
  left_join(weather_conditions, by = c("weather_condition_id" = "weather_condition_id")) %>%
  rename(weather_condition_description = description) %>%
  left_join(wind_directions, by = c("wind_direction_id" = "wind_direction_id")) %>%
  rename(wind_direction_description = description) %>%
  left_join(pollution_levels, by = c("pollution_id" = "pollution_id")) %>%
  rename(pollution_level_description = level) %>%
  left_join(pollen_levels, by = c("pollen_id" = "pollen_id")) %>%
  rename(pollen_level_description = level) %>%
  left_join(uv_index_levels, by = c("uv_index_id" = "uv_index_id")) %>%
  rename(uv_index_level_description = level)

# Remove rows where humidity, pressure, or wind_speed are 0
data <- data %>%
  filter(humidity != 0, pressure != 0, wind_speed != 0)







# Find the minimum low_temperature for each day
daily_min_low_temp <- data %>%
  group_by(date) %>%
  summarise(min_low_temp = min(low_temperature[low_temperature > 0], na.rm = TRUE))

# Join the daily minimum low temperature back to the original data
data <- data %>%
  left_join(daily_min_low_temp, by = "date")

# Replace 0 and NA low_temperature values with the daily minimum low temperature
data <- data %>%
  mutate(
    low_temperature = ifelse(is.na(low_temperature) | low_temperature == 0, min_low_temp, low_temperature)
  )


# Remove the temporary min_low_temp column
data <- data %>%
  select(-min_low_temp)









# Check the cleaned data

summary(data)
```




```{r, echo =FALSE}
# Histogram for high temperatures
ggplot(data, aes(x = high_temperature)) + 
  geom_histogram(binwidth = 2, fill = "blue", color = "white") +
  labs(title = "Distribution of High Temperatures", x = "High Temperature (°C)", y = "Frequency") +
  theme_minimal()
```






```{r, echo=FALSE}
# Histogram for wind speed
ggplot(data, aes(x = wind_speed)) + 
  geom_histogram(binwidth = 1, fill = "orange", color = "white") +
  labs(title = "Distribution of Wind Speed", x = "Wind Speed (km/h)", y = "Frequency") +
  theme_minimal()
```


```{r, echo=FALSE}
# Scatter plot of high temperature vs. wind speed
ggplot(data, aes(x = high_temperature, y = wind_speed)) +
  geom_point(color = "blue") +
  labs(title = "High Temperature vs. Wind Speed", x = "High Temperature (°C)", y = "Wind Speed (km/h)") +
  theme_minimal()
```


```{r, echo=FALSE}
# Convert 'weather_condition_description' to a factor for better plotting
data$weather_condition_description <- factor(data$weather_condition_description)

# Box plot of high temperature by weather condition with a cleaner design
ggplot(data, aes(x = weather_condition_description, y = high_temperature, fill = weather_condition_description)) +
  geom_boxplot(outlier.size = 2, outlier.colour = "red") +  # Highlight outliers
  labs(title = "High Temperature by Weather Condition",
       x = "Weather Description",
       y = "High Temperature (°C)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
        legend.position = "none")  # Remove legend if it’s clear from the plot
```



```{r, echo=FALSE}
# Time series plot of high temperature
ggplot(data, aes(x = time_of_search, y = high_temperature)) +
  geom_line(color = "blue") +
  labs(title = "High Temperature Over Time", x = "Time", y = "High Temperature (°C)") +
  theme_minimal()
```






```{r, echo=FALSE}
# Correlation matrix plot (if you have multiple numeric columns)
numeric_data <- data %>% select(high_temperature, low_temperature, current_temperature, wind_speed, humidity, pressure)
correlation_matrix <- cor(numeric_data, use = "complete.obs")
corrplot(correlation_matrix, method = "circle")
```






```{r, echo=FALSE}
# Density plot of current temperature
ggplot(data, aes(x = current_temperature, fill = "Current Temperature")) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Current Temperature", x = "Current Temperature (°C)", y = "Density") +
  theme_minimal()
```





```{r, echo=FALSE}
# Convert time to date
data$date <- as.Date(data$time_of_search)

# Convert tide heights to numeric
data$low_tide_morning_height <- as.numeric(as.character(data$low_tide_morning_height))
data$high_tide_morning_height <- as.numeric(as.character(data$high_tide_morning_height))
data$low_tide_evening_height <- as.numeric(as.character(data$low_tide_evening_height))
data$high_tide_evening_height <- as.numeric(as.character(data$high_tide_evening_height))

# Get the daily maximum and minimum tide heights
daily_tides <- data %>%
  group_by(date) %>%
  summarise(
    max_tide_height = max(c(low_tide_morning_height, high_tide_morning_height, low_tide_evening_height, high_tide_evening_height), na.rm = TRUE),
    min_tide_height = min(c(low_tide_morning_height, high_tide_morning_height, low_tide_evening_height, high_tide_evening_height), na.rm = TRUE)
  )

# Plot the daily maximum and minimum tide heights
ggplot(daily_tides, aes(x = date)) +
  geom_line(aes(y = max_tide_height, color = "Maximum Tide Height"), linewidth = 1) +
  geom_line(aes(y = min_tide_height, color = "Minimum Tide Height"), linewidth = 1) +
  labs(title = "Daily Maximum and Minimum Tide Heights",
       x = "Date",
       y = "Tide Height (m)",
       color = "Tide Type") +
  scale_color_manual(values = c("Maximum Tide Height" = "red", 
                                "Minimum Tide Height" = "blue")) +
  theme_minimal() +
  theme(legend.position = "bottom")

```





```{r, echo=FALSE}
# Clean data: Convert columns to numeric
data$high_temperature <- as.numeric(gsub("[^0-9.]", "", as.character(data$high_temperature)))
data$wind_speed <- as.numeric(gsub("[^0-9.]", "", as.character(data$wind_speed)))
data$humidity <- as.numeric(gsub("[^0-9.]", "", as.character(data$humidity)))

# Bubble plot of high temperature vs. wind speed with bubble size representing humidity
ggplot(data, aes(x = high_temperature, y = wind_speed, size = humidity, color = humidity)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(1, 15)) +
  scale_color_viridis_c() +
  labs(title = "Bubble Plot of High Temperature vs. Wind Speed",
       x = "High Temperature (°C)",
       y = "Wind Speed (km/h)",
       size = "Humidity (%)",
       color = "Humidity (%)") +
  theme_minimal()

```








```{r, echo =FALSE, warning=FALSE}
# Ensure sunrise, sunset are properly formatted as times and date is in Date format
data$sunrise <- as.POSIXct(data$sunrise, format = "%H:%M", tz = "UTC")
data$sunset <- as.POSIXct(data$sunset, format = "%H:%M", tz = "UTC")

data$date <- as.Date(data$date)
ggplot(data, aes(x = date)) +
  geom_line(aes(y = sunrise), size = 1.2, color = "blue") +
  geom_line(aes(y = sunset), size = 1.2, color = "red") +
  geom_point(aes(y = sunrise), size = 2, color = "blue") +
  geom_point(aes(y = sunset), size = 2, color = "red") +
  labs(title = "Sunrise and Sunset Times Over Time",
       x = "Date",
       y = NULL) +  # Set y label to NULL to remove label text
  scale_y_datetime(date_labels = "%H:%M",  # Format y-axis to show time only
                   breaks = scales::date_breaks("2 hours")) +  # Break every 2 hours
  theme_minimal()


```



```{r bubble_plot_filtered, echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary libraries
library(tidyverse)
library(scales)  # For custom color scales

# Filter data to exclude chance of precipitation values less than 1%
filtered_data <- data %>%
  filter(chance_of_precipitation >= 1)

# Bubble plot of chance of precipitation with fixed small size bubbles and a color scheme from red to blue
ggplot(filtered_data, aes(x = time_of_search, y = chance_of_precipitation, color = current_temperature)) +
  geom_point(size = 2, alpha = 0.7) +  # Set a fixed small size for all bubbles
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient from blue (low temp) to red (high temp)
  labs(title = "Bubble Plot of Chance of Precipitation Over Time",
       x = "Time",
       y = "Chance of Precipitation (%)",
       color = "Current Temperature (°C)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```


